Follow-up Samples

### 2017-01-18 submission/manuscript_outline_20161024.Rmd

Interestingly, when only previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.  This data suggests that the changes that may be important in the bacterial microbiome after surgery are not necessarily any one specific bacterium, a clear depression of CRC related bacteria, nor any addition of new or depressed bacterium but rather that a shift of the bacterial community as a whole occurs with the existing members of this microbiome.

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or initial and follow up samples in order to confirm that the community was what was responding or changing due to surgery.  We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to comensal bacteria.  With only the lesion model having a single OTU from previously associated cancer bacteria (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT both models (lesion and initial and follow up) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].  Confirming the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  One potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  An alternative reason would be that even if they were reduced in relative abundance they were still present.  Another potential explanation is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  It is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  Finally, it is also possible that all three of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be stated that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or initial and follow up).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our Random Forest based model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the fecal DNA test [@cotter_long-term_2016] could help improve our overall AUC.  Another limitation is that this study drew heavily from those with caucasian ancestry.  The results may not be immediately representative of those with either Asian or African ancestry.  Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

Another interesting outcome was that Within figure 3 the initial and follow up model showed better test AUC results then the training set AUC.  This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be overfit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.

>>>>

Interestingly, when only previously associated cancer bacteria were investigated only `r length(rownames(filter(main_crc_bugs_summary, crc_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` had a significant decrease in relative abundance between initial and follow up for the carcinoma group and `r length(rownames(filter(main_crc_bugs_summary, adn_BH < 0.05)))` out of the `r length(rownames(main_crc_bugs_summary))` were significant for the adenoma group.  This data suggests that the changes that may be important in the bacterial microbiome after surgery are not necessarily any one specific bacterium, a clear depression of CRC related bacteria, or any addition of new or depressed bacterium.  Instead it may be due to a shift of the bacterial community as a whole between existing members of the host's microbiome.

We next created a model that incorporated FIT and the bacterial microbiome to either be able to classify lesions (adenoma or carcinoma) or before samples in order to confirm that the community was what was responding or changing due to surgery.  We found that the OTUs that made up the most important variables to the model overwhelmingly belonged to comensal bacteria.  With only the lesion model having a single OTU from a previously associated cancer bacterium (*Porphyromonas asaccharolytica*).  Using only these important OTUs and FIT both models (lesion and before sample) had a significantly decreased positive probability of either lesion or being an initial sample on follow up [Figure 4 & S4].  Further confirmation of the importance of the changes of comensal bacteria to these classifications was that a total of `r length(common_vars_summary$otu)` OTUs were common to both models and the vast majority belonged to regular residents of our gut community.

@@

Curiously, we observed that the typical CRC associated bacteria were not predictive within our models.  There are a number of reasons why this may have occured.  First, one potential explanation is that even with surgery and a shift of the bacterial community these specific bacteria still persisted within the colon.  Second, the bacteria even if they were reduced in relative abundance they were still present within the gut.  Third, is that they were not present in enough individuals to be able to classify those with and without disease with a high degree of accuracy.  Fourth, is that it is possible that our Random Forest models were able to get the same information from measures such as FIT or other OTUs.  Finally, it is also possible that all three of these potential explanations could have played a role.  Regardless, our observations would suggest that an individual's resident bacteria have a large role to play in disease initiation and could change in a way that allows predictive models to lower the positive probability of a lesion after surgery [Figure 4].  It should be noted that our study does not argue against the importance of these CRC associated bacteria in the pathogenesis of disease but rather that the models do not utilize these specific bacteria for classification purposes (lesion or before sample).  In fact, it is possible that these CRC associated bacteria are important in the transition from adenoma to carcinoma and would be one explanation as to why in our data we not only see high initial relative abundances, in certain individuals, but also large decreases in relative abundance in those with carcinoma but not in those with adenoma after surgery [Figure 2].

One limitation of our study is that we do not know whether individuals who were still classified as positive by the lesion model eventually had a subsequent CRC diagnosis.  This information would help to strengthen the case for our Random Forest based model to have kept a number of individuals above the cutoff threshold even though at follow up they were diagnosed as no longer having a lesion.  Another limitation is that we do not know if adding modern tests such as the stool DNA test [@cotter_long-term_2016] could help improve our overall AUC.  Another limitation is that this study drew heavily from those with caucasian ancestry.  The results may not be immediately representative of those with either Asian or African ancestry.  Finally, although our training and test set are relatively large we still run the risk of overfitting or having a model that may not be immediately extrapolateable to other populations.  We've done our best to safeguard against this by not only running 10-fold cross validation but also having over 100 different 80/20 splits to try and mimic the type of variation that might be expected to occur.

Another interesting outcome was that Within figure 3 the before sample model showed better test AUC results then the training set AUC.  This may have occured because the training AUC that was determined from 20 repeated 10 fold cross validation removed samples at random and did not take into account that they were matched samples.  Another potential reason is that the model itself may be overfit since the total number of samples was not that large.  However, the lesion model did not suffer from these discrepencies and similar conclusions can be drawn solely from this model.  Regardless, further independent studies will need to be carried out to verify our findings since not only are we dealing with feces, which could be very different than the communities present on the actual tissue, but also are dealing with correlations that may not be representative of the true pathogensis of disease.
